---
title: "class 5: data viz with ggplot"
author: "Chris Yi (A16849780)"
format: pdf
editor: visual
---

## Intro to ggplot

There are many graphics systems in R (ways to make plots and figures). These include "base" R plots. Today we will focus mostly on the **ggplot2** package. 

Let's start with a plot of a simple in-built dataset called `cars`. 
```{r}
cars
```

```{r}
plot(cars)
```

Let's see how we can make this figure using **ggplot**. First we need to install this package on the computer. To install any R package, use function `install.packages()`. 

> I will run `install.packages("ggplot2) in my R console not quarto document!

Before I can use any functions from add on packages I need to load the package from my "library()" with the `library(ggplot2)` call. 

```{r}
library(ggplot2)
ggplot(cars)
```

All ggplot figures have at least 3 things(called layers). These include:

- **data** (the input dataset I want to plot from) 
- **aes** (the aesthetic mapping of the data to my plot)
- **geoms** (the geom_point(), geom_line() etc. that I want to draw)

```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point()
```



Let's add a line to show the relationship here:

```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth(method="lm") + theme_bw() + labs(title="My First GGPLOT")
```

> Q1 Which geometric layer should be used to create scatter plots in ggplot2?

geom(point)

##Gene expression Figure

The code to read the dataset

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> How many genes are in this dataset?

```{r}
nrow(genes)
```

A first plot of this dataset

```{r}
main_p <- ggplot(genes) + aes(x=Condition1, y=Condition2, col=State) + geom_point() + scale_color_manual( values=c("blue", "gray", "red" )) + theme_bw() + labs(title="Gene Expression changes upon drug treatment", x="Control Expression", y="Drug Presence Expession")
main_p
```

> Use the table() function on the State column of this data.frame to find out how many ‘up’ regulated genes there are. What is your answer?

```{r}
table(genes$State)
```

> Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset?

```{r}
round( table(genes$State)/nrow(genes), 4)
```

```{r}
n.tot <- nrow(genes)
vals <- table(genes$State)

vals.percent <- vals/n.tot * 100
round(vals.percent, 2)
```
> Q. Use the colnames() function and the ncol() function on the genes data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find?
```{r}
ncol(genes)
```

```{r}
p <- ggplot(genes) + aes(x=Condition1, y=Condition2, col=State) + geom_point() 
```

```{r}
p + labs(title="Hello") + theme_minimal()
```

>Q.For which phases is data visualization important in our scientific workflows?
All of the above: communication of results, exploratory data analysis (EDA), detection of outliers

>Q. True or False? The ggplot2 package comes already installed with R?
False

>Q. Which plot types are typically NOT used to compare distributions of numeric variables?
Network graphs

>Q. Which statement about data visualization with ggplot2 is incorrect?
ggplot2 is the only way to create plots in R

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
p1 <- ggplot(gapminder) + aes(x=gdpPercap, y=lifeExp) + geom_point()
p1
```

```{r}
library(patchwork)
(main_p | p1) 
```

```{r}
library(dplyr)
gapminder_2007 <- gapminder %>% filter(year==2007)
```

```{r}
p2 <- ggplot(gapminder_2007) + aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) + geom_point(alpha=0.20)
p2                                                                        
```

```{r}
library(patchwork)
(main_p / p2)
```

